Aligning sequence reads, clone sequences and assembly contigs with BWA-MEM

### 2013-03-14 bwamem.tex
a little more wording improvement

\author[Li]{Heng Li$^{1,}$\footnote{to whom correspondence should be addressed}}

\address{$^1$Broad Institute, 7 Cambridge Center, Cambridge, MA 02142, USA}

@@

\begin{abstract}
\section{Summary:} BWA-MEM is a new alignment algorithm for aligning sequence
reads or long query sequences against a large reference genome such as human.
It automatically chooses between local and end-to-end alignments, supports
paired-end reads and performs split alignment. The algorithm is robust to
sequencing errors and applicable to a wide range of sequence lengths from 70bp
to a couple of megabases. For short-read mapping, BWA-MEM shows better
performance than many state-of-art read aligners to date.

@@

\vspace*{-1em}
\begin{methods}

@@

\subsubsection{Seeding and re-seeding} BWA-MEM follows the canonical
seed-and-extend paradigm. It initially seeds an alignment with supermaximal
exact matches (SMEMs) using an algorithm we found previously~\citep[Algorithm
5]{Li:2012fk}, which essentially finds at each query position the longest exact
match covering the position. However, occasionally the true alignment may not
contain any SMEMs.  To reduce mismappings caused by missing seeds, we introduce
re-seeding. Suppose we have a SMEM of length $l$ with $k$ occurrences in the
reference genome. If $l$ is too large (over 28bp by default), we then re-seed
with exact matches that cover the middle base of the SMEM and occur at least
$k+1$ times in the genome. Such seeds can be found by requiring a minimum
occurrence in the original SMEM algorithm.

\subsubsection{Chaining and chain filtering} We call a group of seeds that are
colinear and close to each other as a \emph{chain}. We chain the seeds while
seeding. After that, we filter out short chains that are largely contained in a
long chain and are much worse than the long chain (by default, both 50\% and
38bp shorter than the long chain). Note that chain filtering aims to reduce
unsuccessful seed extension at a later step. Each chain may not always
correspond to a final hit. The chains detected here do not need to be accurate.

\subsubsection{Seed extension} We rank a seed by the length of the chain it
belongs to and then by the seed length in the chain. For each seed in the
ranked list, from the best to the worst, we drop the seed if it is already
contained in an alignment found before, or extend the seed with a banded
affine-gap-penalty dynamic programming (DP) if it potentially leads to a new
alignment.

@@

Secondly, while extending a seed, BWA-MEM tries to keep track of the best
extension score reaching the end of the query sequence. If the difference
between the best score reaching the end and the best local alignment score is
below a threshold, the local alignment will be rejected even if it has a higher
score. BWA-MEM uses this strategy to automatically choose between local and
end-to-end alignments. It may allow true substitutions towards the end of a
read and thus reduces reference bias, while avoids introducing excessive
mismatches and gaps which may happen when we force an end-to-end alignment for
a read bridging a break point associated with a long insertion or deletion
(indel) or a structural variation or a misassembly.

\begin{figure}[tb]
\centering
\includegraphics[width=0.4\textwidth]{eval/alnroc-se}\\
\includegraphics[width=0.4\textwidth]{eval/alnroc-pe}
\caption{Percent mapped reads as a function of the false alignment rate under
different mapping quality cutoff. $10^6$ pairs of 101bp reads are simulated
from the human reference genome using wgsim (http://bit.ly/wgsim2) with 1.5\%
substitution errors and 0.2\% indel variants. The insert size follows a normal
distribution $N(500,50^2)$. The reads are aligned back to the genome either as
single end (SE; top panel) or as paired end (PE; bottom panel). GEM is
configured to allow up to 5 gaps and to output suboptimal alignments (option
`--e5 --m5 --s1' for SE and `--e5 --m5 --s1 --pb' for PE). Its mapping quality
is estimated with a BWA-like algorithm. Other mappers are run essentially with
the default setting except specifying the insert size distribution. The run
time on a single CPU core is shown in the parantheses.}\label{fig:eval}
\end{figure}.

@@

BWA-MEM is a fast and accurate aligner for sequence reads and is one of the few
that work well for both short reads and long sequences up to a few megabases.
Technically, it is possible to make BWA-MEM even faster for long sequences by
using SSE2-based banded DP and by restricting DP to regions not covered by a
long exact match.

>>>>

\author[Li]{Heng Li}

\address{Broad Institute of Harvard and MIT, 7 Cambridge Center, Cambridge, MA 02142, USA}

@@

\begin{abstract}
\section{Summary:} BWA-MEM is a new alignment algorithm for aligning sequence
reads or long query sequences against a large reference genome such as human.
It automatically chooses between local and end-to-end alignments, supports
paired-end reads and performs split alignment. The algorithm is robust to
sequencing errors and applicable to a wide range of sequence lengths from 70bp
to a few megabases. For short-read mapping, BWA-MEM shows better performance
than many state-of-art read aligners to date.

@@

\begin{methods}

@@

\subsubsection{Seeding and re-seeding} BWA-MEM follows the canonical
seed-and-extend paradigm. It initially seeds an alignment with supermaximal
exact matches (SMEMs) using an algorithm we found previously~\citep[Algorithm
5]{Li:2012fk}, which essentially finds at each query position the longest exact
match covering the position. However, occasionally the true alignment may not
contain any SMEMs.  To reduce mismappings caused by missing seeds, we introduce
re-seeding. Suppose we have a SMEM of length $l$ with $k$ occurrences in the
reference genome. If $l$ is too large (over 28bp by default), we re-seed
with the longest exact matches that cover the middle base of the SMEM and occur
at least $k+1$ times in the genome. Such seeds can be found by requiring a
minimum occurrence in the original SMEM algorithm.

\subsubsection{Chaining and chain filtering} We call a group of seeds that are
colinear and close to each other as a \emph{chain}. We chain the seeds while
seeding. After that, we filter out short chains that are largely contained in a
long chain and are much worse than the long chain (by default, both 50\% and
38bp shorter than the long chain). Chain filtering aims to reduce unsuccessful
seed extension at a later step. Each chain may not always correspond to a final
hit. Chains detected here do not need to be accurate.

\subsubsection{Seed extension} We rank a seed by the length of the chain it
belongs to and then by the seed length. For each seed in the ranked list, from
the best to the worst, we drop the seed if it is already contained in an
alignment found before, or extend the seed with a banded affine-gap-penalty
dynamic programming (DP) if it potentially leads to a new alignment.

@@

Secondly, while extending a seed, BWA-MEM tries to keep track of the best
extension score reaching the end of the query sequence. If the difference
between the best score reaching the end and the best local alignment score is
below a threshold, the local alignment will be rejected even if it has a higher
score. BWA-MEM uses this strategy to automatically choose between local and
end-to-end alignments. It may align through true variations towards the end of
a read and thus reduces reference bias, while avoids introducing excessive
mismatches and gaps which may happen when we force a chimeric read into an
end-to-end alignment.

\begin{figure}[tb]
\centering
\includegraphics[width=0.4\textwidth]{eval/alnroc-se}\\
\includegraphics[width=0.4\textwidth]{eval/alnroc-pe}
\caption{Percent mapped reads as a function of the false alignment rate under
different mapping quality cutoff. $10^6$ pairs of 101bp reads are simulated
from the human reference genome using wgsim (http://bit.ly/wgsim2) with 1.5\%
substitution errors and 0.2\% indel variants. The insert size follows a normal
distribution $N(500,50^2)$. The reads are aligned back to the genome either as
single end (SE; top panel) or as paired end (PE; bottom panel). GEM is
configured to allow up to 5 gaps and to output suboptimal alignments (option
`--e5 --m5 --s1' for SE and `--e5 --m5 --s1 --pb' for PE). Its mapping quality
is estimated with a BWA-like algorithm. Other mappers are run essentially with
the default setting except for specifying the insert size distribution. The CPU
time in seconds is reported in the parantheses.}\label{fig:eval}
\end{figure}.

@@

BWA-MEM is a fast and accurate aligner for sequence reads and is one of the few
that work well for both short reads and long sequences up to a few megabases.
Technically, it is possible to make BWA-MEM even faster for long sequences by
using SSE2-based banded DP and by restricting DP to regions not covered by a
long exact match. Seeding is the bottleneck for short sequences, while banded
DP is the bottleneck for long sequences.
